Defining Greatness

### 2024-12-01 paper/model_card.qmd
Add files via upload

**Input Data**:  
- Dataset: Player-level NFL game statistics from `merged_player_game.parquet`.  
- Features: 16 numerical features, including completions, attempts, passing yards, passing touchdowns, interceptions, and rushing statistics.  
- Training Data: Filtered observations of Tom Brady's career game statistics.  
- Prediction Data: Patrick Mahomes' per-game averages, projected over the next 200 games.

@@

**Evaluation Metrics**: The model's fit and prediction accuracy were evaluated using the following metrics:  
- Adjusted R² values for multivariate regression.  
- Mean squared error (MSE) for the fitted model.

**Performance on Training Data (Tom Brady)**:  
- The model successfully captured trends in Tom Brady’s career statistics, showing high adjusted R² values for most features.  
- No missing data points were included in the training subset.

**Projection Validation**:  
- Model predictions were cross-referenced with Mahomes' historical averages.  
- Assumes statistical continuity in performance.

@@

1. **Contextual Limitations**: The model assumes that Patrick Mahomes' future performance mirrors his historical averages and does not account for potential changes in playing style, team composition, or career disruptions.

>>>>

**Input Data**:  
- Dataset: Player-level NFL game statistics from `merged_player_game.parquet`.  
- Features: Includes completions, attempts, passing yards, passing touchdowns, interceptions, and rushing statistics.  
- Training Data: Filtered observations of Tom Brady's career game statistics.  
- Prediction Data: Patrick Mahomes' per-game averages, projected over the next 200 games.

@@

**Evaluation Metrics**: The model's fit and prediction accuracy were evaluated using RMSE following

**Performance on Training Data (Tom Brady)**:  
- The model somewhat successfully captured trends in Tom Brady’s career statistics

@@

1. **Contextual Limitations**: The model assumes that Patrick Mahomes' future performance mirrors Tom Brady's and does not account for potential changes in playing style, team composition, or career disruptions.

### 2024-12-01 paper/paper.qmd
Add files via upload

The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Arrow [@citearrow], Knitr [@citeknitr], and Testthat [@citett] packages. The primary dataset included QB-specific data from every NFL game played since 2000, encompassing over 55 columns detailing various player performance metrics. Additional data sources were used for ESPN’s Quarterback Rating (QBR), which encapsulates QB efficiency across multiple dimensions of gameplay.

@@

## Variables of Interest
Most of the variables used in this analysis are relatively straight forward. Passing yards is the vertical distance a passing play gains. For example, if a quarterback throws a pass and the receiver catches 10 yards down the field and runs for another 10 yards before being tackled or going out of bounds, that will count as 20 passing yards for the Quarterback. Rushing yards are the number of yards a player runs with the ball past the line of scrimmage - if the ball is caught as a pass, it does not count towards rushing yards. Passing touchdowns are simply a passing play that turns into a touchdown and rushing touchdowns are a rush that turns into a touchdown. Interceptions are passes made by a quarterback and caught by the other team.

@@

## Model Overview

We designed a predictive model to estimate Patrick Mahomes' cumulative career statistics over an additional 200 games. The model is trained on historical game data from Tom Brady as a benchmark and Patrick Mahomes' existing career data to generate predictions for key performance metrics.

### Mathematical Notation

The model is a multivariate linear regression defined as:

$$
\hat{y}_i = \beta_0 + \sum_{j=1}^p \beta_j x_{ij} + \epsilon_i
$$

@@

- $\hat{y}_i$: Predicted value of the response variable (e.g., total passing yards).  
- $\beta_0$: Intercept term.  
- $\beta_j$: Coefficients for predictors $x_{ij}$, representing the influence of the $j$-th feature on the response variable.  
- $x_{ij}$: Observed value of the $j$-th predictor for the $i$-th observation.  
- $\epsilon_i$: Residual error term, assumed to be normally distributed with mean 0.

We estimate $\beta_0$ and $\beta_j$ using Ordinary Least Squares (OLS) on Tom Brady's historical data, as his career provides a well-documented trajectory for long-term performance.

### Variables and Justification

The predictors ($x_j$) included in the model are chosen based on their relevance to a quarterback's performance:

1. **Passing Metrics**:  
   - `completions`, `attempts`, `passing_yards`, `passing_tds`, `interceptions`.  
   - These metrics directly influence a quarterback’s overall contribution to team success.

2. **Rushing Metrics**:  
   - `carries`, `rushing_yards`, `rushing_tds`.  
   - While not the primary focus, rushing performance is an insightful metric for dual-threat quarterbacks like Mahomes.

These features ensure the model captures both primary and auxiliary aspects of quarterback performance.

### Model Assumptions

1. **Linearity**: The relationship between predictors and response variables is linear.  
2. **Independence**: Residuals are independent across observations.  
3. **Homoscedasticity**: Residuals have constant variance.  
4. **Normality**: Residuals are normally distributed.  
5. **Stationarity**: Mahomes' performance trajectory will follow trends observed in Brady's career.

### Software and Implementation

The model was implemented using R, leveraging the `lm` function from the `stats` package for regression. The dataset was preprocessed using `tidyverse`, and predictions were calculated for an additional 200 games using the coefficients estimated from Brady's data.

### Validation and Diagnostics

1. **Train-Test Split**: The Brady dataset was split into training (80%) and testing (20%) sets. The model was trained on the training set and validated on the testing set.  
2. **Error Metrics**:  
   - **Root Mean Squared Error (RMSE)**: Measures the model's predictive accuracy.  
   - **Mean Absolute Error (MAE)**: Evaluates average prediction error.  
3. **Residual Analysis**:  
   - Residual plots confirmed the assumptions of homoscedasticity and normality.  
4. **Out-of-Sample Testing**: Predicted Mahomes' statistics on his observed data and compared with actual results to ensure alignment.

@@

1. **Decision Trees**:  
   - Strengths: Captures non-linear relationships.  
   - Weaknesses: Tends to overfit without pruning; less interpretable than linear regression.

2. **Bayesian Regression**:  
   - Strengths: Allows incorporation of priors, producing probabilistic predictions.  
   - Weaknesses: Increased complexity and computational requirements.

3. **Final Choice**:  
   - Linear regression was chosen for its balance of simplicity, interpretability, and performance. It aligns with the assumption that a quarterback’s performance trends over time can be captured linearly.

### Limitations

1. **Career Longevity**:  
   - The model assumes Mahomes will continue performing at a consistent level for 200 games. This may not account for potential injuries, performance decline, or external factors. This is already a very generous assumption, considering only two quarterbacks in NFL history have played more than 324 career games, but because Brady has played 381 games, it is important for our analysis to assume that Mahomes will be able to sustain success. 
2. **Sample Bias**:  
   - Relying on Brady's career as a benchmark may introduce bias, as it assumes Mahomes will follow a similar trajectory.  
3. **Feature Engineering**:  
   - Excluding contextual factors like team strength or play style may limit predictive accuracy.

### Conclusion

The linear regression model provides a robust yet interpretable framework for predicting Mahomes' lifetime stats. It is grounded in historical data and validated through residual analysis and error metrics. While limitations exist, the model effectively captures the essence of Mahomes' expected performance trajectory based on Brady's historical patterns.

# Results
 [@tbl-brady-mahomes-career-stats] compares Tom Brady’s career statistics with Patrick Mahomes’ projected lifetime statistics. Tom Brady, with 8,954 completions and 13,973 attempts, has significantly more career completions and attempts than Mahomes, who is projected to finish with 7,756.72 completions and 11,902.61 attempts. Brady’s total passing yards of 102,616 far exceed Mahomes’ projected 89,548.38, highlighting Brady’s longevity and consistent performance. Similarly, Brady has thrown 738 passing touchdowns, whereas Mahomes is projected to finish with 660.38. Brady also has more interceptions (252 compared to Mahomes’ 213.59). These differences underscore Brady’s long career and high volume of passing, which has contributed to his unrivaled career achievements.

@@

When comparing rushing statistics, Mahomes stands out as the more mobile quarterback. While Brady’s rushing stats are modest, with 1,256 rushing yards and 35 rushing touchdowns, Mahomes is projected to amass 3,282.88 rushing yards and 36.28 rushing touchdowns, showing a clear advantage in rushing ability. Additionally, Mahomes is expected to have 319.90 rushing first downs, surpassing Brady’s 266. This emphasizes Mahomes’ dual-threat capability, which contributes to his versatility as a quarterback.

@@

The analysis of per-game statistics shows a clear edge for Patrick Mahomes over Tom Brady across nearly every metric. Mahomes completes slightly more passes per game (23.94) than Brady (23.50), despite attempting a comparable number of throws (36.74 for Mahomes vs. 36.67 for Brady). This small difference results in a higher completion percentage for Mahomes at 65.17%, compared to Brady’s 64.08%, indicating a marginal but noteworthy advantage in passing efficiency. Mahomes also outpaces Brady in passing yards per game, averaging 276.38 to Brady’s 269.33, further emphasizing his ability to generate more offense through the air.

In terms of scoring, Mahomes demonstrates a superior passing touchdown rate, averaging 2.04 touchdowns per game compared to Brady’s 1.94. Both players maintain an impressive level of ball security, with nearly identical interception rates per game (0.66).

The most significant disparity appears in rushing statistics, where Mahomes outshines Brady by a wide margin. Mahomes averages 10.13 rushing yards per game, more than triple Brady’s 3.30, and his higher number of carries per game (2.85 vs. 2.12) underscores his ability to contribute as a dual-threat quarterback. Furthermore, Mahomes’ rushing touchdown rate (0.11 per game) surpasses Brady’s 0.09, further illustrating his dynamic impact on the ground.

>>>>

The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Caret [@citecaret], Arrow [@citearrow], Knitr [@citeknitr], and Testthat [@citett] packages. The primary dataset included QB-specific data from every NFL game played since 2000, encompassing over 55 columns detailing various player performance metrics. Additional data sources were used for ESPN’s Quarterback Rating (QBR), which encapsulates QB efficiency across multiple dimensions of gameplay.

@@

## Variables of Interest
Most of the variables used in this analysis are relatively straight forward. Passing yards is the vertical distance a passing play gains. For example, if a quarterback throws a pass and the receiver catches 10 yards down the field and runs for another 10 yards before being tackled or going out of bounds, that will count as 20 passing yards for the Quarterback. Rushing yards are the number of yards a player runs with the ball past the line of scrimmage - if the ball is caught as a pass, it does not count towards rushing yards. Passing touchdowns are simply a passing play that turns into a touchdown and rushing touchdowns are a rush that turns into a touchdown. Interceptions are passes made by a quarterback and caught by the other team. Carries are how many rush attempts the QB has.

@@

The goal of this analysis is to predict Patrick Mahomes' lifetime career statistics over 200 additional games using a regression model. This model uses data from Tom Brady's career as a proxy for player performance and is designed to forecast Mahomes' future performance across key metrics, such as completions, attempts, passing yards, rushing yards, and touchdowns.

## Data

The dataset used in this analysis is a combination of game-level statistics for two players: Tom Brady and Patrick Mahomes. The data includes multiple variables that track each player's performance across different games. The key features used in this model are:

- **completions**: The number of successful passes thrown.
- **attempts**: The total number of pass attempts.
- **passing_yards**: The total yards gained from passing attempts.
- **passing_tds**: The total number of passing touchdowns.
- **interceptions**: The total number of interceptions thrown.
- **carries**: The number of rushing attempts.
- **rushing_yards**: The total rushing yards gained.
- **rushing_tds**: The total number of rushing touchdowns.

The model used in this analysis is a multiple linear regression model. We choose linear regression because it is a simple yet effective approach for modeling continuous outcomes and provides easy interpretability.

### Equation for the Model

The model predicts the following variables for Patrick Mahomes:

- \( \text{completions}, \text{attempts}, \text{passing yards}, \text{passing tds}, \text{interceptions}, \text{carries}, \text{rushing yards}, \text{rushing tds} \)

The general form of the multiple linear regression model is:

$$
Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_n X_n + \epsilon
$$

@@

- $Y$ represents the outcome variable (e.g., completions, attempts, passing yards, etc.).
- $X_1, X_2, \dots, X_n$ are the predictor variables (game statistics such as completions, attempts, etc.).
- $\beta_0$ is the intercept.
- $\beta_1, \beta_2, \dots, \beta_n$ are the coefficients for each predictor.
- $\epsilon$ is the error term.

### Training and Validation

We trained the model using data from Tom Brady's career, specifically using the following steps:

1. **Data Preparation**: We selected relevant features and ensured there were no missing values in the training dataset.
2. **Model Creation**: A multiple linear regression model was created using the `lm()` function in R. The model predicts the selected statistics for each game.
3. **Training/Test Split**: The dataset was split into training (80%) and testing (20%) sets to evaluate model performance.
4. **Prediction**: The trained model was used to predict Mahomes' statistics for the next 200 games.

We also calculated the Root Mean Squared Error (RMSE) for each predicted variable in the test set as an evaluation metric. RMSE is a commonly used metric to measure the accuracy of regression models, and it is calculated as:

$$
RMSE = \sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2}
$$

- $y_i$ is the actual value of the statistic.

- $\hat{y}_i$ is the predicted value.

- $n$ is the number of test samples.

The RMSE values for each variable in the test set are as follows:

completions: 6.01,
attempts: 9.57,
passing yards: 75.90,
passing touchdowns: 1.28,
interceptions: 0.80,
carries: 1.72,
rushing yards: 4.70 and
rushing touchdowns: 0.15

These RMSE values indicate that the model has some level of predictive accuracy but also considerable error in certain variables, such as passing yards. The RMSE is higher than expected, which can be attributed to the limited dataset size used in this model. The dataset contains a relatively small number of games, which leads to higher variability and uncertainty in the predictions.

@@

1. **Decision Trees**:  
   - Strengths: Captures non-linear relationships.  
   - Weaknesses: Tends to overfit without pruning; less interpretable than linear regression, and still had high error.

2. **Bayesian Regression**:  
   - Strengths: Allows incorporation of priors, producing probabilistic predictions.  
   - Weaknesses: Increased complexity and computational requirements and also had high error.

3. **Final Choice**:  
   - Linear regression was chosen for its balance of simplicity, interpretability, and performance.

## Limitiations

While the RMSE values suggest some error in the predictions, this is expected due to the limited amount of available data. The performance of the model could likely improve with a larger and more balanced dataset that includes a broader range of game conditions. Additionally, while we used Tom Brady’s statistics as a proxy for Mahomes, differences in their playing styles and team dynamics could also contribute to the prediction error.

The linear regression model assumes that the relationships between the predictors and the outcomes are linear, which may not always be the case in real-world scenarios. Further exploration of alternative models, such as non-linear regression or machine learning techniques like decision trees, could be considered for improving predictions.

## Conclusion

This model provides a reasonable estimate of Patrick Mahomes’ lifetime career statistics over 200 additional games based on Tom Brady’s historical data. Although the RMSE values suggest room for improvement, the model serves as a useful starting point for projecting Mahomes’ future performance. Future work could include expanding the dataset, exploring non-linear modeling techniques, and accounting for player-specific factors that might differ between Brady and Mahomes.

# Results
 [@tbl-brady-mahomes-career-stats] compares Tom Brady’s career statistics with Patrick Mahomes’ projected lifetime statistics. Tom Brady, with 8,954 completions and 13,973 attempts, has significantly more career completions and attempts than Mahomes, who is projected to finish with 7,763 completions and 11,925 attempts. Brady’s total passing yards of 102,616 far exceed Mahomes’ projected 89,160, highlighting Brady’s longevity and consistent performance. Similarly, Brady has thrown 738 passing touchdowns, whereas Mahomes is projected to finish with 654. Brady also has more interceptions (252 compared to Mahomes’ 216). These differences underscore Brady’s long career and high volume of passing, which has contributed to his unrivaled career achievements.

@@

When comparing rushing statistics, Mahomes stands out as the more mobile quarterback. While Brady’s rushing stats are modest, with 1,256 rushing yards and 35 rushing touchdowns, Mahomes is projected to amass 3,348 rushing yards and 40 rushing touchdowns, showing a clear advantage in rushing ability. This emphasizes Mahomes’ dual-threat capability, which contributes to his versatility as a quarterback.

@@

The analysis of per-game statistics shows a clear edge for Patrick Mahomes over Tom Brady across nearly every metric. Mahomes completes slightly more passes per game (23.96) than Brady (23.50), despite attempting a comparable number of throws (36.81 for Mahomes vs. 36.67 for Brady). This small difference results in a higher completion percentage (defined as the number of completions divided by the number of attempts) for Mahomes at 65.10%, compared to Brady’s 64.08%, indicating a marginal but noteworthy advantage in passing efficiency. Mahomes also outpaces Brady in passing yards per game, averaging 275.18 to Brady’s 269.33, further emphasizing his ability to generate more offense through the air.

In terms of scoring, Mahomes demonstrates a superior passing touchdown rate, averaging 2.02 touchdowns per game compared to Brady’s 1.94. Both players maintain an impressive level of ball security, with nearly identical interception rates per game (0.66 for Brady to 0.67 for Mahomes).

The most significant disparity appears in rushing statistics, where Mahomes outshines Brady by a wide margin. Mahomes averages 10.33 rushing yards per game, more than triple Brady’s 3.30, and his higher number of carries per game (2.87 vs. 2.12) underscores his ability to contribute as a dual-threat quarterback. Furthermore, Mahomes’ rushing touchdown rate (0.12 per game) surpasses Brady’s 0.09, further illustrating his dynamic impact on the ground.
